---
title: "Applied Micro PSET 3"
author: "Matthew Zhao"
date: "2023-01-16"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(haven)
library(stargazer)
library(xtable)
library(lmtest)
library(sandwich)
library(broom)
options(scipen=999)
```

## Question 1

```{r}
df1 <- read_stata('data/Y1.dta')
df2 <- read_stata('data/Y2.dta')
```

### a)

One of the key motivators that the authors cite is that in developing nations, there has been a large push in making primary school universal and specifically getting students enrolled in primary school. However, the literature reveals that while we are able to get students to attend primary school, this has had little impact on actual learning, with students typically many grade levels behind where they should be in terms of core competencies e.g. literacy and numeracy. 

A survey of past studies reveals that additional resources e.g. smaller classroom size or more textbooks, has little effect for the average student but some effect for students who were above average. As a result, the authors in this paper look to seek if supporting weaker students with remedial education, i.e. providing education that is catered towards their actual level of knowledge, and more broadly if resources which target/adjust to a student's actual level of understanding, would produce gains in literacy and numeracy. 

### b)

https://evalf21.classes.andrewheiss.com/example/standard-errors/

```{r}
# cleaning and standardizing data
year1df <- df1 %>%
  select(!c(starts_with('mid'),contains('norm'))) %>%
  filter(!(is.na(pre_verb) | is.na(pre_math) | is.na(post_verb) | is.na(post_math)))

y1_mean_pre_verb_ctrl <- mean((year1df %>% filter(bal == 0))$pre_verb)
y1_sd_pre_verb_ctrl <- sd((year1df %>% filter(bal == 0))$pre_verb)
y1_mean_pre_math_ctrl <- mean((year1df %>% filter(bal == 0))$pre_math)
y1_sd_pre_math_ctrl <- sd((year1df %>% filter(bal == 0))$pre_math)
y1_mean_pre_tot_ctrl <- mean((year1df %>% filter(bal == 0))$pre_tot)
y1_sd_pre_tot_ctrl <- sd((year1df %>% filter(bal == 0))$pre_tot)

y1.cleaned <- year1df %>% 
  mutate(pre_verbnorm = (pre_verb - y1_mean_pre_verb_ctrl)/y1_sd_pre_verb_ctrl,
         pre_mathnorm = (pre_math - y1_mean_pre_math_ctrl)/y1_sd_pre_math_ctrl,
         pre_totnorm = (pre_tot - y1_mean_pre_tot_ctrl)/y1_sd_pre_tot_ctrl,
         post_verbnorm = (post_verb - y1_mean_pre_verb_ctrl)/y1_sd_pre_verb_ctrl,
         post_mathnorm = (post_math - y1_mean_pre_math_ctrl)/y1_sd_pre_math_ctrl,
         post_totnorm = (post_tot - y1_mean_pre_tot_ctrl)/y1_sd_pre_tot_ctrl
         )
```

```{r, warning=F}
# regression balance test
bal_test1_mlr <- lm(bal ~ pre_totnorm + male, data = y1.cleaned)
bal_test1_pretotnorm <- lm(bal ~ pre_totnorm, data = y1.cleaned)
bal_test1_male <- lm(bal ~ male, data = y1.cleaned)
models <- list(bal_test1_pretotnorm,bal_test1_male,bal_test1_mlr)
stargazer(models,type='text',digits=3,
          title='Table 1 - Balance Test')
```

We find that the coefficients are insignificantly different from 0 with the implication that the groups are likely balanced. Standard errors for this analysis should be clustered since treatment is randomized at the school level, hence the SEs should be clustered at the school level. 

### c)

```{r, warning=F}
# table II reproduction
t2_df <- y1.cleaned %>% 
  select(c(schoolid,bal,contains('norm'))) %>%
  pivot_longer(cols = c(contains('norm')),names_to = 'score_cat',values_to='score') %>%
  mutate(bal_cat = ifelse(bal == 1,'Treatment','Comparison')) %>%
  group_by(bal_cat,score_cat) %>%
  summarise(means = )
  pivot_wider(names_from = bal_cat, values_from = score)

# pre_verb_mean <- lm(bal ~ pre_totnorm, data = y1.cleaned)
# bal_test1_male <- lm(bal ~ male, data = y1.cleaned)
# models <- list(bal_test1_pretotnorm,bal_test1_male,bal_test1_mlr)
# stargazer(models,type='text',digits=3,
#           title='Table 1 - Balance Test')
```

### d)



### e)



### f)



### g)

```{r}
case1 <- read_stata('data/case1.dta')
```



### h)



### i)



### j)



### k)



### l)

```{r}
case2 <- read_stata('data/case2.dta')
```



### m)



### n)



### o)



### p)



### q)



### r)



### s)

```{r}
case3 <- read_stata('data/case3.dta')
```



### t)



### u)



### v)



### w)



### x)



### y)



### z)



## Problem 2



### a)



### b)



### c)



### d)





